Placeholder for article 3

### 2025-02-18 analysis/paper/paper.qmd
Macbook morning session upload: Complete assemblage description and tables moved to supplementary. Main paper tables simplified with categories collapsed for an easier read. tbl- chunks added tbl-subcap to present table groups (assemblage and retouched) in a single plot.

# In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
solutrean_WIP <- solutrean_WIP |> 
  mutate(Class = case_when( # This alteration is being done on the original class column
    Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
    TRUE ~ Class)) |> 
  mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
    Class == "Blade" | Class == "Bladelet" ~ "ElongBlank", # Cluster blades with bladelets. The distinction can come later through a mutate and the measurements
    Class == "Core front" | Class == "Core tablet" | Class == "Core trimming elem." | Class == "Crested piece" ~ "CorePrepProd", # Cluster all preparation products
    Class == "Fragment" ~ "Shatter",
    Class == "Flake distal frag." | Class == "Flake mesial frag." | Class == "Flake prox. frag." ~ "FlakeFrag", # Cluster flake frags into a single category
    Class == "Blade distal frag." | Class == "Blade mesial frag." | Class == "Blade prox. frag." 
    | Class == "Bladelet distal frag." | Class == "Bladelet mesial frag." | Class == "Bladelet prox. frag." ~ "ElongBlankFrag", # Cluster elong frags into a single category
    TRUE ~ Class)) |> 
  mutate(COREPREPPROD = case_when( # Creating a column for the types of core preparation and maintenance products (COREPREPPROD)
    Class == "Core front" ~ "Core front",
    Class == "Core tablet" ~ "Core tablet",
    Class == "Core trimming elem." ~ "Core trim",
    Class == "Crested piece" ~ "Crested piece")) |> 
  mutate(PIECECOMPLETENESS = case_when( # Creating a column for the portions present in flake or elongblank fragments to match other DBs
    Class == "Flake distal frag." | Class == "Blade distal frag." | Class == "Bladelet distal frag." ~ "Distal",
    Class == "Flake mesial frag." | Class == "Blade mesial frag." | Class == "Bladelet mesial frag." ~ "Mesial",
    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) |> 
  rename(PLATFORMTYPE = Platform_type) |> # Renaming column
  mutate(PLATFORMTYPE = case_when( # Renaming variables to match correspondent variables in the other DBs
    PLATFORMTYPE == "Multifacetted" ~ "Faceted",
    PLATFORMTYPE == "Unfacetted" ~ "Plain",
    TRUE ~ PLATFORMTYPE)) |>
  mutate(Type = case_when( # Translating retouched types from Zilhão 1997 to simplified typologies
    Type == "92b" ~ "Retouched piece fragment",
    Type == "92a" ~ "Retouched blank",
    Type == "76" ~ "Splintered piece",
    Type == "75" ~ "Denticulate",
    Type == "74" ~ "Notch",
    Type == "66" ~ "Retouched blade",
    Type == "61" ~ "Truncation",
    Type == "28" ~ "Burin",
    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
    TRUE ~ Type)) |>
  rename(LEVEL = Layer, UNIT = Unit, CORTEX = Percent_of_cortex, CORTEXLOCATION = Loc_of_cortex, THICKNESS = Thickness,
         MAXWIDTH = Width, LENGTH = Length, COREPLATFORM = Core_platfform, SCARCOUNT = N_of_Scars, COREABANDON = Core_abandon,
         RETOUCHEDPIECETYPOLOGY = Type)  # Renaming column

@@

# Tables with class by raw material type (n and %)
## For Gravettian
gravettian_general <- phase_dataset |>
  filter(Level %in% c(6, 7)) |>
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |>
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Proto-Solutrean (4E and top of 5)
proto_general <- phase_dataset |> 
  filter(Level %in% c("4E", "4E/5")) |> 
  filter(CLASS != "Chip") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Solutrean
solutrean_general <- solutrean_dataset |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL", "INDET"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

@@

# Solutrean table
## Table preparation
solutrean_retouch <- solutrean_dataset |>
    mutate(CLASS = case_when(
    CLASS == "Retouched piece" ~ "RetouchedPiece",
    CLASS == "Retouched piece fragment" ~ "RetouchedPieceFrag",
    TRUE ~ CLASS)) |>
  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
  filter(RETOUCHEDPIECETYPOLOGY != "") |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    RMTYPE == "TYPE 11" ~ "T11",
    TRUE ~ RMTYPE))
## Janitor tabyl for final table with n and %
solutrean_retouch_tab <- solutrean_retouch |>
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |>
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T5, T6, T7, T8, T11, INDET, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

# Proto-solutrean table
## Table preparation
proto_retouch <- phase_dataset |>
  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
  filter(Level %in% c("4E", "4E/5")) |>
  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    RMTYPE == "TYPE 11" ~ "T11",
    TRUE ~ RMTYPE))
## Janitor tabyl for final table with n and %
proto_retouch_tab <- proto_retouch |>
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |> 
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, TL, INDET, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

# Gravettian table
## Table preparation
gravettian_retouch <- phase_dataset |> 
  filter(CLASS %in% c("RetouchedPiece", "RetouchedPieceFrag")) |>
  filter(Level %in% c(6, 7)) |> 
  select(RMTYPE, RETOUCHEDPIECETYPOLOGY) |> 
  dplyr::group_by(RMTYPE) |>
  dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" ~ "T1",
    RMTYPE == "TYPE 2" ~ "T2",
    RMTYPE == "TYPE 3" ~ "T3",
    RMTYPE == "TYPE 4" ~ "T4",
    RMTYPE == "TYPE 5" ~ "T5",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    TRUE ~ RMTYPE))
## Janitor tabyl for final table with n and %
gravettian_retouched_tab <- gravettian_retouch |> 
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |>
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, T1, T2, T3, T4, T5, T6, T7, T8, T9, T10, TL, INDET, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

@@

```{r}
#| label: tbl-general-gravettian
#| tbl-cap: Number and percentage of technotypological classes by raw material type from the Gravettian assemblage (levels 6 and 7) of the Terrace area of Vale Boi.

kable(gravettian_general)

@@

```{r}
#| label: tbl-retouch-gravettian
#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian assemblage (levels 6 and 7) of the Terrace area.

kable(gravettian_retouched_tab)

>>>>

# In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
solutrean_WIP <- solutrean_WIP |> 
  mutate(Class = case_when( # This alteration is being done on the original class column
    Retouched == "Yes" ~ "RetouchedPiece", # Whenever retouch is present, class is now Retouched piece
    TRUE ~ Class)) |> 
  mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
    Class == "Blade" | Class == "Bladelet" ~ "ElongBlank", # Cluster blades with bladelets. The distinction can come later through a mutate and the measurements
    Class == "Core front" | Class == "Core tablet" | Class == "Core trimming elem." | Class == "Crested piece" ~ "CorePrepProd", # Cluster all preparation products
    Class == "Fragment" ~ "Shatter",
    Class == "Flake distal frag." | Class == "Flake mesial frag." | Class == "Flake prox. frag." ~ "FlakeFrag", # Cluster flake frags into a single category
    Class == "Blade distal frag." | Class == "Blade mesial frag." | Class == "Blade prox. frag." 
    | Class == "Bladelet distal frag." | Class == "Bladelet mesial frag." | Class == "Bladelet prox. frag." ~ "ElongBlankFrag", # Cluster elong frags into a single category
    TRUE ~ Class)) |> 
  mutate(COREPREPPROD = case_when( # Creating a column for the types of core preparation and maintenance products (COREPREPPROD)
    Class == "Core front" ~ "Core front",
    Class == "Core tablet" ~ "Core tablet",
    Class == "Core trimming elem." ~ "Core trim",
    Class == "Crested piece" ~ "Crested piece")) |> 
  mutate(PIECECOMPLETENESS = case_when( # Creating a column for the portions present in flake or elongblank fragments to match other DBs
    Class == "Flake distal frag." | Class == "Blade distal frag." | Class == "Bladelet distal frag." ~ "Distal",
    Class == "Flake mesial frag." | Class == "Blade mesial frag." | Class == "Bladelet mesial frag." ~ "Mesial",
    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) |> 
  rename(PLATFORMTYPE = Platform_type) |> # Renaming column
  mutate(PLATFORMTYPE = case_when( # Renaming variables to match correspondent variables in the other DBs
    PLATFORMTYPE == "Multifacetted" ~ "Faceted",
    PLATFORMTYPE == "Unfacetted" ~ "Plain",
    TRUE ~ PLATFORMTYPE)) |>
  mutate(Type = case_when( # Translating retouched types from Zilhão 1997 to simplified typologies
    Type == "92b" ~ "Retouched piece fragment",
    Type == "92a" ~ "Retouched blank",
    Type == "76" ~ "Splintered piece",
    Type == "75" ~ "Denticulate",
    Type == "74" ~ "Notch",
    Type == "66" ~ "Retouched blade",
    Type == "61" ~ "Truncation",
    Type == "28" ~ "Burin",
    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
    TRUE ~ Type)) |>
  rename(LEVEL = Layer, UNIT = Unit, CORTEX = Percent_of_cortex, CORTEXLOCATION = Loc_of_cortex, THICKNESS = Thickness,
         MAXWIDTH = Width, LENGTH = Length, COREPLATFORM = Core_platfform, SCARCOUNT = N_of_Scars, COREABANDON = Core_abandon,
         RETOUCHEDPIECETYPOLOGY = Type)  # Renaming column

@@

# Tables with class by raw material type (n and %)
##
general_table <- phase_dataset |> 
    dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    TRUE ~ RMTYPE)) |> 
  dplyr::mutate(CLASS = case_when(
    CLASS == "BlankFrag" ~ "Blank fragment",
    CLASS == "BurinSpall" ~ "Burin spall",
    CLASS == "CoreFrag" ~ "Core fragment",
    CLASS == "CorePreparProd" ~ "Core prep product",
    CLASS == "RetouchedPiece" ~ "Retouched",
    CLASS == "RetouchedPieceFrag" ~ "Retouched fragment",
    TRUE ~ CLASS)) |> 
  dplyr::mutate(RETOUCHEDPIECETYPOLOGY = case_when(
    RETOUCHEDPIECETYPOLOGY == "Retouched blade" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet" | RETOUCHEDPIECETYPOLOGY == "Retouched bladelet fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake" | RETOUCHEDPIECETYPOLOGY == "Retouched flake fragment" ~ "Retouched blank",
    RETOUCHEDPIECETYPOLOGY == "Splintered piece fragment" ~ "Splintered piece",
    TRUE ~ RETOUCHEDPIECETYPOLOGY))

general_table_solutrean <- solutrean_dataset |> 
   dplyr::mutate(RMTYPE = case_when(
    RMTYPE == "TYPE 1" | RMTYPE == "TYPE 2" | RMTYPE == "TYPE 3" | RMTYPE == "TYPE 4" | RMTYPE == "TYPE 5" ~ "Local",
    RMTYPE == "TYPE 6" ~ "T6",
    RMTYPE == "TYPE 7" ~ "T7",
    RMTYPE == "TYPE 8" ~ "T8",
    RMTYPE == "TYPE 9" ~ "T9",
    RMTYPE == "TYPE 10" ~ "T10",
    RMTYPE == "TYPE 11" ~ "T11",
    TRUE ~ RMTYPE)) |> 
  dplyr::mutate(CLASS = case_when(
    CLASS == "ElongBlank" | CLASS == "Flake" ~ "Blank",
    CLASS == "ElongBlankFrag" | CLASS == "FlakeFrag" ~ "Blank fragment",
    CLASS == "Thinning flake" ~ "Blank",
    CLASS == "CorePrepProd" ~ "Core prep product",
    CLASS == "RetouchedPiece" | CLASS == "Retouched piece" ~ "Retouched",
    CLASS == "Retouched piece fragment" ~ "Retouched fragment",
    TRUE ~ CLASS))

## For Gravettian
gravettian_general <- general_table |>
  filter(Level %in% c(6, 7)) |>
  filter(RMTYPE != "INDET") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |>
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Proto-Solutrean (4E and top of 5)
proto_general <- general_table |> 
  filter(Level %in% c("4E", "4E/5")) |> 
  filter(CLASS != "Chip") |> 
  filter(RMTYPE != "INDET") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
                                        "T8", "T9", "T10", "TL"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

## For Solutrean
solutrean_general <- general_table_solutrean |> 
  filter(RMTYPE != "INDET") |> 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Local", "T6","T7", 
                                        "T8", "T9", "T10", "T11", "TL"))) |> 
  tabyl(CLASS, RMTYPE) |> 
  adorn_totals(c("row", "col")) |> 
  adorn_percentages("col") |> 
  adorn_pct_formatting(2) |> 
  adorn_ns("front") |> 
  mutate(across(where(is.character), ~ gsub("^0   \\(0\\.00%\\)$", " ", .)))

@@

# Solutrean table
## Table preparation
solutrean_retouch <- general_table_solutrean |>
  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
  filter(RMTYPE != "INDET") |> 
  filter(RETOUCHEDPIECETYPOLOGY != "") |> 
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |>
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T11, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

# Proto-solutrean table
## Table preparation
proto_retouch <- general_table |>
  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
  filter(Level %in% c("4E", "4E/5")) |>
  filter(!RMTYPE == "INDET") |> 
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |> 
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, TL, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

# Gravettian table
## Table preparation
gravettian_retouch <- general_table |> 
  filter(CLASS %in% c("Retouched", "Retouched fragment")) |>
  filter(Level %in% c(6, 7)) |> 
  filter(!RMTYPE == "INDET") |> 
  tabyl(RETOUCHEDPIECETYPOLOGY, RMTYPE) |>
  adorn_totals(c("row", "col")) |>
  mutate(across(where(is.numeric), ~ ifelse(. == 0, " ", .))) |>
  select(RETOUCHEDPIECETYPOLOGY, Local, T6, T7, T8, T9, T10, TL, Total) |> # Ordering values
  rename(Typology = RETOUCHEDPIECETYPOLOGY)

@@

```{r}
#| label: tbl-general
#| tbl-cap: Number and percentage of technotypological classes by raw material type of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages.
#| tbl-subcap: 
#| - "Number and percentage of technotypological classes by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
#| - "Number and percentage of technotypological classes by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
#| - "Number and percentage of technotypological classes by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
#| layout-nrow: 4

kable(head(gravettian_general))
kable(head(proto_general))
kable(head(solutrean_general))

@@

```{r}
#| label: tbl-retouch
#| tbl-cap: Number and percentage of retouched tools by raw material of the Gravettian (a), Proto-Solutrean (b) and Solutrean (c) assemblages. Retouched blank typology includes flakes, blades and bladelets and respective fragments.
#| tbl-subcap: 
#| - "Number and percentage of retouched tools by raw material from the Gravettian assemblage (levels 6 and 7) of the Terrace area."
#| - "Number and percentage of retouched tools by raw material from the Proto-Solutrean assemblage (levels 4E and 5) of the Terrace area."
#| - "Number and percentage of retouched tools by raw material from the Solutrean assemblage (levels A-C) of the Shelter area."
#| layout-nrow: 4

kable(head(gravettian_retouched))
kable(head(proto_retouch))
kable(head(solutrean_retouch))
